Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-03-31 paper/metamonads.tex
a better draft of the CPS section

The inlining is not perfect however. We would like that our code
expression to look like
\begin{code}
 fun arr n ->
   let pv = ref (0,arr.(0)) in ... the rest of fpca ...
   in let i = fst !pv in ...
\end{code}

We would like that the inlining of |.~(fpca abs_float .<arr>. .<n>.)|
generated us the |let pv = ref ... | statement. But this seems to
quite a challenge. We wish the resulting code to look
|let pv = ... in body| where |pv = ...| part come from |fpca| and
|body| come from |upv|. At first blush, this seems impossible. A code
expression may contain only complete expressions, such as
|let v =... in ...|. A code expression cannot contain only the binding part of
|let|, which is not an expression per se.

The solution: CPS. And here where I go to sleep.

>>>>

The generator |fpca| shows one particular way of searching for pivot.
In the domain of exact rational numbers, for example, we would be
satisfied with the first non-zero element as a pivot, and so we would
write a different pivot-searching generator. The body of |upv| moves
the found pivot to the ``canonical'' location. Again, depending on the
container (a dense matrix, a sparse matrix, etc), there are different
ways of doing that. So, we would like to be able to ``compose'' pivot
generating function with the swap generating function. We would like
to write something like this

\begin{code}
let fpca absf arr n pv = .<
   for i = 1 to .~n-1 do
     if absf (.~arr).(i) > snd ! (.~pv) then
        (.~pv) := (i, absf (.~arr).(i))
   done>.
let swapper arr n pv = .<
   let i = fst !(.~pv) in
   if not (i = 0) then
      let t = (.~arr).(0) in
      begin (.~arr).(0) <- (.~arr).(i); (.~arr).(i) <- t end>.
let seqv a b c = .<begin .~a; .~b; .~c end>.
let pv_swap absf arr n =
   let pv = .<ref (0,(.~arr).(0))>. in
   seqv (fpca absf arr n pv)
        (swapper arr n pv)
        arr
let top =  .<fun arr n -> .~(pv_swap abs_float .<arr>. .<n>.)>.
\end{code}

The function |pv_swap| shows our ideal -- composing code gerenators
|fpca| and |swapper|, using a higher-order combinator `seqv'
(sequence of code fragments). We lifted the pivot declaration out of
old version of |fpca| so that both |fpca| and |swapper| could be
parameterized by the pivot (and |fpca| and similar function will deal
only with pivot determination). |top| puts it all together. The above
code represents the pattern we will be following for the rest of the
paper.

And it \emph{almost} works. It is type correct.  When we look at the
generated code for |top|, we notice the `(ref (0, arr_1.(0)))' is
inlined wherever pivot was expected. For example, the swapping part of
the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in
...|. That is certainly not what we want. We meant for pivot searching
and swapping parts of the code to communicate via a shared mutable
state, variable |pv|. In the generated code, each piece of code gets
its own copy of |pv|. Rather then inline (ref (0, arr_1.(0)))
everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
and inlined |pv|. Incidentally, even in a pure functional code we
often wish to generate such a |let| statement so we can compute a
complex expression only once and use it value in several other
places. Again, we would rather not leave the job of detecting such
common subexpressions to the compiler (because in the complex code the
compiler may fail to detect the common subexpressions). The problem
with generating such |let| expressions is that we wish our geenrator
to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
of the |let|, which we would fill in later. In languages which
generate code in the string form this is easy. We generate the string
|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
representting the body, wcich would give us the complete |let|
statement. In MetaOCaml, we can't generate incomplete expressions, for
a good reason.

To solve the problem of writing a generator for a |let| expression 
when the generated body is not yet known, we use CPS.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
And now, a better -- and shorter -- write-up.

The essence of our approach is assembling the code of the program by combining
instances of various aspects together. In the case of Gaussian
Elimination, the aspects are pivot determination, swapping rows and
columns, determinant tracking, etc. A primitive code generator yields a
code expression 
A combinator takes (complete and well-typed) code expressions combines
them into a composite expression (which is too statically guaranteed
to be well-formed and well-typed).

For example, the following is probably the simplest code generators:
\begin{code}
let one = .<1>.
\end{code}

(in the floating-poiint instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code fragments (perhaps say one more sentence about the brackets).
The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes, which are
...

We can see the result in
\begin{code}
let simplest_code =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

MetaOCaml helpfully prints out the result of the latter, as
|.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that 'gen x y' looks like a combinator; the body of
'gen x y' involves several apoplications of the function 'plus'. When
we look at the generated code however, we see no traces of function
'plus' and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the above code: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
simplest_code is not the function on integers -- it's a code
expression.

To see the benefit of code generation, we notice that we can easily
parameterize our code:

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

and use it to generate code that operates on integers, floating pint
numbers or boolean -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a bollean function.

We can see the separation of concerns, implementation of domain
operations.

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

We see that expression `(plus y one)' appears twice. That may be quite
a complex computation (depending on the implementation of `plus'), and
so we would rather not do it twice. We may be tempted to rely on the
compiler's built-in common-subexpression elimination algorithm, which
is a typical part of the optimizer. We would rather not relied on the
sufficiently smart compiler: when the generated code is very complex,
the compiler may overlook common subexpressions (or the subexpression
occur in such an imperative context that the compiler might not be
able to decide if lifting the subexpression is sound. So, being
conservative, the optimizer will leave the duplicates). We may attempt
to do
\begin{code}
let param_code1 plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

However, examinine the code for |param_code1 plus one|
shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
we might have saved the computation on the code generation
stage. However, the generated code still exhibits the
duplication (which makes the code larger and may take longer to
execute, if the optimizer fails to eliminate the duplication).
So, we need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| so we could use
that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
But that seems impossible because |letgen exp| needs to generate
the let expression .<let t = exp in body>. but |letgen| doesn't have
the body yet. The body needs that temporary identifier .<t>. that will
be generated by |letgen| itself. Certainly |letgen| cannot generate
only a part of a let-expression. All generated expressions in
MetaOCaml are well-formed and complete.

The key is to use CPS. It was first pointed out by XXX in the context
of partial evaluation, and extensively used by (FFT) for code
generation.

\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
\end{code}

Now, |param_code2 plus one| gives us the desired code
|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
